%\newm{
\color{blue}
\textbf{Changes}

\begin{itemize}
	\item Removed text related to examiners Section ``Acknowledgement''.
	\item Abstract:
		\begin{itemize}
			\item Changed ``advancements'' to ``advances''
			\item Added a sentence introducing what intrinsic and post-hoc methods are.
			\item Minor fix
		\end{itemize}
	\item \autoref{sec:achiml}:
		\begin{itemize}
			\item Paraphrased the last paragraph.
			\item Add a sentence to mention existing governmental regulations and
				recommendations on AI.
				\item Minor fix.
		\end{itemize}
	\item \autoref{sec:risexai}:
		\begin{itemize}
			\item Clarified which references are surveys and which ones are papers.
		\end{itemize}
	\item \autoref{sec:motiv}:
		\begin{itemize}
			\item Mentioned Formal XAI~(FXAI) is an model-specific approach. (FXAI
				$\neq$ model-specific approach because some heuristic approaches are
				also model specific, such as TreeSHAP.)
			\item Added more context in contribution 3
			\item Minor fix.
		\end{itemize}
	\item \autoref{sec:rq}:
		\begin{itemize}
			\item Adjusted section counters.
			\item Updated publications
			\item Added a sentence in contribution 1 to describe our approach.
			\item Minor fix.
		\end{itemize}
	\item Section ``Publications during enrolment''
		\begin{itemize}
			\item Text and attached paper: Replaced the arxiv version of our paper ``Anytime Approximate Formal
				Feature Attribution.'' by the SAT conference paper.
			\item Text and attached paper: Replaced the Just-in-time explainer paper
				with the latest version. 
		\end{itemize}

	\item \autoref{sec:ml}
		\begin{itemize}
			\item \autoref{sec:nn} Added a math expression with brief activations
				functions introduced
			\item Minor fix
		\end{itemize}
	\item \autoref{sec:xai}
		\begin{itemize}
			\item \autoref{sec:malimit}: Added a sentence mentioning issue in exact
				Shapley values.
			\item \autoref{sec:fxai}: Added the definition of Minimal hitting set
				duality.
			\item Minor fix.
		\end{itemize}
	\item \autoref{chap:jair21}: Added a reference for a recent study.
	\item \autoref{chap:ffa}: Minor fix.
	\item \autoref{chap:conc}
		\begin{itemize}
			\item Minor fix.
		\end{itemize}
	
\end{itemize}
%}
\color{black}
