% https://www.monash.edu/graduate-research/examination/publication
% https://www.monash.edu/rlo/graduate-research-writing/write-the-thesis
% https://www.monash.edu/rlo/graduate-research-writing/write-the-thesis/writing-the-thesis-chapters/structuring-a-long-text
%\ack{
%\addtocontents{toc}{}  % Add a gap in the Contents, for aesthetics

\section*{Acknowledgements}
\addtotoc{Acknowledgements}

% beginning
The journey of pursuing my Ph.D. is both challenging and rewarding in my life.
%
Without the guidance, encouragement, and support from my supervisors, colleagues, and family, 
this achievement is not attainable.

% thank supervisors
I am deeply grateful for the supervision provided by Prof. Peter J. Stuckey and 
Dr. Alexey Ignatiev throughout my research journey.
%
Their guidance and support have been invaluable,  and their thoughtful manner to science
has a significant factor in driving my work.
%
I express my gratitude to them for granting me the scholarship from their funding,
which has played a crucial role in supporting my Ph.D. studies.
%
Furthermore, I am sincerely thankful to them for acquainting me with the exciting
realm of explainable AI.
%
This exploration of explainable AI has not only broadened my perspective
but has also fueled my enthusiasm for AI research.
%
I am fortunate to have supervisors who actively engage in every one of my research projects, 
providing constant guidance and assistance.
%
Moreover, I highly value the flexibility they grant me to investigate a range of interesting
problems.
%
To both my supervisors, I extend my sincere gratitude for the dedication you have 
shown in guiding me, ultimately shaping me through the Ph.D journey.


% thanks panels and collaborators
I would like to express my gratitude to all members of the panel committee, 
Assoc. Prof. Arun Konagurthu, Assoc. Prof. Guido Tack, and Prof. Geoff Webb, 
for their invaluable feedback and support throughout my candidature.
%
This thesis has significantly benefited from the suggestions provided by the examiners 
XXX and XXX.
%
%Although my thesis committee members underwent some changes, I am truly thankful for their constructive feedback, insightful suggestions, and encouragement during this process. 
%
Additionally, I extend my appreciation to my collaborators 
Dr. Pierre Le Bodic, Prof. Joao Marques-Silva,
Dr. Nina Narodytska, Dr. Chakkrit Tantithamthavorn, Michael Fu, and Prof. Graham Farr.
%
Their dedication and hard work on my research projects are deeply appreciated.
%
Their invaluable assistance and guidance are significant, without which I 
would not be able to complete these research works.
%
% Your contributions have been critical in shaping the outcomes of my work, and I am truly grateful for your support and collaboration.


% thanks colleagues
I will like to thank my fellow researchers and colleagues at the OPTIMA lab for 
fostering a stimulating and supportive environment that has contributed to 
my intellectual development.
%
Also, I like to express my gratitude to Dr. Bojie Shen, Hendrik Bierlee, Kelvin Davis, Jiarong Fan,
Shizhe Zhao, Dr. Allen Zhong, Dr. Terrence Mak, Sushmita Paul, Dr. Jip Dekker, and other members for
engaging in numerous insightful research discussions with me, and for the interesting
micro talks we enjoyed on Fridays.
%
Their encouragement and support have been crucial in inspiring me to
successfully complete my research journey.

% Thank friends and family
Finally, I deeply appreciate my family for their love, unconditional support, and 
continuous encouragement during my Ph.D. journey.
%
Their belief in my capabilities has consistently provided me with motivation
and strength even in the toughest time. 
%
I am sincerely thankful to them for being instrumental in my journey.
%
Without their support, I would not have reached this point.
