% https://www.monash.edu/graduate-research/examination/publication
% https://www.monash.edu/rlo/graduate-research-writing/write-the-thesis
% https://www.monash.edu/rlo/graduate-research-writing/write-the-thesis/writing-the-thesis-chapters/structuring-a-long-text
%\Publicationslist{

%\addtocontents{toc}{}  % Add a gap  \addtocontents{toc}{\vspace{1em}}  in the Contents, for aesthetics

\section*{Publications during enrolment}
\addtotoc{Publications during enrolment}

The full citation of work declared in this ``thesis by publication'' compilation is as follows:

{\large

1. Jinqiang Yu, Alexey Ignatiev, Peter J. Stuckey, and Pierre Le Bodic. Learning optimal decision
sets and lists with sat. \emph{Journal of Artificial Intelligence Research,} 72, 1251-1279, 2021.

2. Jinqiang Yu, Alexey Ignatiev, Peter J. Stuckey, Nina Narodytska, and Joao Marques-Silva.
Eliminating the impossible, whatever remains must be true: On extracting and applying background
knowledge in the context of formal explanations. \emph{In Proceedings of the AAAI Conference on Artificial
Intelligence,} vol. 37, pp. 4123-4131, 2023.

3. Jinqiang Yu, Alexey Ignatiev, and Peter J. Stuckey. From formal boosted tree explanations to
interpretable rule sets. \emph{In 29th International Conference on Principles and Practice of
	Constraint Programming,} vol. 280, pp.38:1-38:21, 2023.

4. Jinqiang Yu, Alexey Ignatiev, and Peter J. Stuckey. On Formal Feature Attribution and Its
Approximation. \emph{arXiv preprint arXiv:2307.03380,} 2023.

5. Jinqiang Yu, Graham Farr, Alexey Ignatiev, and Peter J. Stuckey. Anytime Approximate Formal
Feature Attribution. \emph{arXiv preprint arXiv:2312.06973,} 2023.

6.Jinqiang Yu, Michael Fu, Alexey Ignatiev, Chakkrit Tantithamthavorn, and Peter J. Stuckey. A
Formal Explainer for Just-In-Time Defect Predictions. \emph{ACM Transactions on Software Engineering
	and Methodology,} 2024. (\textbf{Minor Revision submitted.})
}
%}
